On the entropy of isothermal dusty gas: application to protoplanetary disks

### 2016-08-04 model.tex
regular update

@@

Finally, we define 
\begin{align}
  Z \equiv \epsilon_0\frac{\Hd}{\Hg} \simeq
  \frac{\Sigma_\mathrm{d}}{\Sigma_\mathrm{g}} 
\end{align}
as a proxy of the ratio between the dust surface density
$\Sigma_\mathrm{d}$ and the gas surface density $\Sigma_\mathrm{g}$.

@@

Writing Eq. \ref{vshear} in terms of the gas density and dust-to-gas
ratio with a power-law temperature profile (Eq. \ref{power_temp}) gives 
\begin{align*}
  &r\frac{\p \Omega^2}{\p z}  =
  \frac{c_s^2(r)}{\left(1+\epsilon\right)^2}\left\{
  \frac{\p\epsilon}{\p r}\frac{\p\ln{\rhog}}{\p z}
  -\frac{\p\epsilon}{\p z}\frac{\p\ln{\rhog}}{\p r}\right.\notag\\
  &\phantom{ r\frac{\p \Omega^2}{\p z}  =
    \frac{c_s^2(r)}{\left(1+\epsilon\right)^2}\left\{\right\} }
  \left. +\frac{q}{r}\left[\frac{\p\epsilon}{\p z} + \left(1+\epsilon\right)\frac{\p\ln{\rhog}}{\p z}\right]
  \right\} 
\end{align*}
The first two terms correspond to vertical shear caused by spatial
variations in the dust-to-gas ratio; while the second term
proportional to $q$ corresponds to vertical shear due to the 
radial temperature gradient.

@@

In this case, the last two integrals in Eq. \ref{int_rel} vanish. Then 
requiring $\sigma^2>0$  gives the Solberg-Hoiland criteria 
for the axisymmtric \emph{stability} of an isothermal gas perfectly 
coupled to dust,  
\begin{align}
  \kappa^2 + c_s^2 \nabla\ln{\rhog}\cdot\nabla\tepsilon &> 0,\label{dusty_solberg1}  \\
  -c_s^2\frac{\p\ln{\rhog}}{\p z}\left(-\kappa^2\frac{\p\tepsilon}{\p
    z} + r\frac{\p\Omega^2}{\p z}\frac{\p \tepsilon}{\p r} \right) & > 0, \label{dusty_solberg2}
\end{align}
%\begin{align}
%  \kappa^2 - \frac{1}{\rho}\nabla P \cdot \nabla s &> 0,\\
%  -\frac{1}{\rho}\frac{\p P}{\p z} \left(\kappa^2 \frac{\p s}{\p z} -
%  r\frac{\p\Omega^2}{\p z}\frac{\p s}{\p r}\right) &>0, 
%\end{align} 
where %$s = \ln{P/\rho}$, 
$\kappa^2 \equiv r^{-3}\p_r\left(r^4\Omega^2\right)$ is the
square of the epicylic frequency, and we assume $\kappa^2>0$. 
Eq. \ref{dusty_solberg1}---\ref{dusty_solberg2} are equivalent to the
Solberg-Hoiland criteria for adiabatic gas if one writes the 
entropy $s=\ln{\left[c_s^2\left(1 - \tepsilon\right)\right]}$.

>>>>

%We comment that

@@

Finally, we define 
\begin{align}
  Z \equiv \epsilon_0\frac{\Hd}{\Hg} \simeq
  \frac{\Sigma_\mathrm{d}}{\Sigma_\mathrm{g}} 
\end{align}
as a measure of the local metalicity. The second approximation holds
for $\epsilon_0\ll1$. 
%\Sigma_\mathrm{d}$ and the gas
%surface density $\Sigma_\mathrm{g}$.

@@

Writing Eq. \ref{vshear} in terms of the gas density and dust-to-gas
ratio with a power-law temperature profile (Eq. \ref{power_temp}) gives 
\begin{align*}
  &r\frac{\p \Omega^2}{\p z}  =
  \frac{c_s^2(r)}{\left(1+\epsilon\right)^2}\left\{
  \frac{\p\epsilon}{\p r}\frac{\p\ln{\rhog}}{\p z}
  -\frac{\p\epsilon}{\p z}\frac{\p\ln{\rhog}}{\p r}\right.\notag\\
  &\phantom{ r\frac{\p \Omega^2}{\p z}  =
    \frac{c_s^2(r)}{\left(1+\epsilon\right)^2}\left\{\right\} }
  \left. -\frac{q}{r}\left[\frac{\p\epsilon}{\p z} + \left(1+\epsilon\right)\frac{\p\ln{\rhog}}{\p z}\right]
  \right\} 
\end{align*}
The first two terms correspond to vertical shear caused by spatial
variations in the dust-to-gas ratio; while the second term
proportional to $q$ corresponds to vertical shear due to the 
radial temperature gradient.

@@

In this case, the last two integrals in Eq. \ref{int_rel} vanish. Then 
requiring $\sigma^2>0$  gives the criteria 
for the axisymmtric \emph{stability} of an isothermal gas perfectly 
coupled to dust,  
%\begin{align}
%  \kappa^2 + c_s^2 \nabla\ln{\rhog}\cdot\nabla\tepsilon &> 0,\label{dusty_solberg1}  \\
%  -c_s^2\frac{\p\ln{\rhog}}{\p z}\left(-\kappa^2\frac{\p\tepsilon}{\p
%    z} + r\frac{\p\Omega^2}{\p z}\frac{\p \tepsilon}{\p r} \right) & > 0, \label{dusty_solberg2}
%\end{align} 
\begin{align}
  \kappa^2 - \frac{1}{\rho}\nabla P \cdot \nabla s &> 0, \label{dusty_solberg1}    \\
  -\frac{1}{\rho}\frac{\p P}{\p z} \left(\kappa^2 \frac{\p s}{\p z} -
  r\frac{\p\Omega^2}{\p z}\frac{\p s}{\p r}\right) &>0, \label{dusty_solberg2}
\end{align} 
where %$s = \ln{P/\rho}$, 
$\kappa^2 \equiv r^{-3}\p_r\left(r^4\Omega^2\right)$ is the 
square of the epicylic frequency, and we assume $\kappa^2>0$.  
Eq. \ref{dusty_solberg1}---\ref{dusty_solberg2} has the identical form
to the Solberg-Hoiland criteria for adiabatic gas dynamics 
\citep[e.g.][]{tassoul78}, except here the entropy $s$ is really a
representation of the dust distribution via $s=\ln{\left[c_s^2\left(1
    - \tepsilon\right)\right]}$.

### 2016-08-04 result.tex
regular update

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_eigenvals_vareps} 
  \caption{Maximum growth rate of the dusty VSI as a function of the
    midplane dust-to-gas ratio $\epsilon_0$ for perturbations with
    different radial wavenumbers $k$. The dust layer thickness is
    fixed to $\Hd\simeq \Hg$. 
    \label{vsi_dust_loading_vareps}
    }
\end{figure}

\subsection{Effect of dust thickness} 
We now vary the dust layer thickness $\Hd$ but hold 
$Z \equiv \epsilon_0 \Hd/\Hg = \text{constant,}$ 
so that $\Sigma_\mathrm{d}/\Sigma_\mathrm{g}$ is
fixed. This then gives the required midplane dust-to-gas ratio.    
Since we will consider thin dust layers, here we use a smaller
domain with $\zmax=2\Hg$ so that $\epsilon$ does not become
too small.

@@

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_vshear_Nz2_fixZ} 
  \caption{Vertical shear rate (solid) compared to vertical buoyancy
    (dotted) in a locally isothermal, dusty disk 
    with dust-to-gas mass ratio $Z=0.03$ and dust thickness $\Hd=0.1\Hg$
    (black) and $\Hd=0.99\Hg$ (blue). 
    \label{compare_vshear_fixZ}
    }
\end{figure}

@@

>>>>

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_eigenvals_vareps2} 
  \caption{Maximum growth rate of the dusty VSI as a function of the
    midplane dust-to-gas ratio $\epsilon_0$ for perturbations with
    different radial wavenumbers $k$. The dust layer thickness is
    fixed to $\Hd\simeq \Hg$. 
    \label{vsi_dust_loading_vareps}
    }
\end{figure}

\subsection{Effect of dust thickness} 
We now vary the dust layer thickness $\Hd$ but fix the metalicity 
$Z \equiv \epsilon_0 \Hd/\Hg = \text{constant}$. This then gives the
required midplane dust-to-gas ratio. Since we will consider thin dust
layers, here we use a smaller 
domain with $\zmax=2\Hg$ so that $\epsilon$ does not become
too small.

@@

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_vshear_Nz2_fixZ} 
  \caption{Vertical shear rate (solid) compared to vertical buoyancy
    (dotted) in a locally isothermal, dusty disk 
    with metalicity $Z=0.03$ and dust thickness $\Hd=0.1\Hg$
    (black) and $\Hd=0.99\Hg$ (blue). 
    \label{compare_vshear_fixZ}
    }
\end{figure}

@@

Fig. \ref{compare_eigenvals_fixZ} shows the maximum VSI growth rates
as a function of the dust layer thickness. As before, we find growth
rates are most affected by the vertical structure of the dust layer
when the perturbation wavenumer is large. VSI growth rates converge as
$\Hd\to 0$. %to dust free values?    
Thus a thin dust layer, however large its associated vertical shear,
does not affect the VSI. The non-monotonic behavior for $\Hd\gtrsim
0.5\Hg$ arises because the vertical buoyancy frequency 
\begin{align*}
N_z^2(H_d;z,Z) \simeq &Z\Hg z^2
\exp{\left(-\frac{z^2}{2\Hg^2}\right)}\OmK^2\notag\\
&\times 
\frac{1}{\Hd}\left(\frac{1}{\Hd^2} -
\frac{1}{\Hg^2}\right)\exp{\left(-\frac{z^2}{2\Hd^2}\right)}  
\end{align*}
is a non-monotonic function of $\Hd$ at fixed $z$. At $z=\Hg$ and $z=2\Hg$
the buoyancy frequency is maximized for $\Hd\simeq0.5\Hg$ and
$\Hd\simeq 0.8\Hg$, respectively. This is consistent with the abscissa
of minima in growth rates in Fig. \ref{compare_eigenvals_fixZ}.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_eigenvals_fixZ} 
  \caption{Maximum VSI growth rate for different perturbation
    wavenumbers $k$ as a function of the dust layer
    thickness $\Hd$ at fixed metalicity $Z=0.03$. 
    \label{compare_eigenvals_fixZ}
    }
\end{figure}

### 2016-08-04 summary.tex
regular update

\subsection{Dusty analogs of other gaseous instabilities} 
%gi
%rwi

@@

\subsection{Thermodynamic interpretation of other dusty instabilities}

>>>>

\subsection{Generalization to locally polytropic disks}
We can extend the dusty/adiabatic gas correspondence to 
other fixed equations of state. As an example, consider the locally
polytropic disk 
\begin{align}
  P = K(r,z)\rhog^{\Gamma}, 
\end{align}
where $K$ is a prescribed function and $\Gamma$ is the constant
adiabatic index. Then eliminating $\tepsilon$ from the dust equation
\ref{dusteq} gives 
\begin{align}
  \frac{D P}{D t} = - \Gamma P\nabla\cdot\bm{v}  + P \bm{v}\cdot\nabla \ln{K}
  + \frac{\Gamma P}{\rhog}\nabla\cdot\left(\tepsilon\tstop\nabla
  P\right).
\end{align}
Thus the dusty gas behaves like a pure gas with adiabatic index
$\Gamma$. The entropy is given by 
\begin{align}
  s = \ln{\left[K(r,z)\left(1 - \tepsilon\right)^\Gamma\right]}.  
\end{align}
When $K$ is constant, the corresponding Solberg-Hoiland criteria for
axisymmetric stability may be obtained from that in standard adiabatic 
hydrodynamics \citep[e.g.][]{tassoul78} by
using this definition of entropy in
Eqs. \ref{dusty_solberg1}---\ref{dusty_solberg2}.

\subsection{Dusty analogs of other gaseous instabilities}

%gi
%rwi

@@

\subsection{Thermodynamic interpretation of dusty instabilities}
